Long-term cloud enhancement events of global solar irradiance over Thessaloniki, Greece

### 2024-08-26 article/article.tex
Commit 2024-08-26 23:11

The data used in this study were recorded at the monitoring site of the
Laboratory of Atmospheric Physics, Aristotle University of Thessaloniki,
in Thessaloniki, Greece (\(40^\circ\,38'\,\)N, \(22^\circ\,57'\,\)E,
\(80\,\)m~a.s.l.). The GHI data were measured with a Kipp~\& Zonen CM-21
pyranometer and cover the period 13~April 1993 to 31~December 2023.
During the study period, the pyranometer was independently calibrated
three times at the Meteorologisches Observatorium Lindenberg, DWD,
verifying that the stability of the instrument's sensitivity was better
than \(0.7\,\%\) relative to the initial calibration by the
manufacturer. For the acquisition of radiometric data, the signal of the
pyranometer was sampled at a rate of \(1\,\text{Hz}\) with the mean and
standard deviation of these samples calculated and recorded every
minute. The measurements were corrected for the zero offset (``dark
signal'' in volts), which was calculated by averaging all measurements
recorded for a period of \(3\,\text{h}\), before (morning) or after
(evening) the Sun reaches an elevation angle of \(-10^\circ\). The
signal was converted to irradiance using the ramped value of the
instrument's sensitivity between subsequent calibrations.

@@

Following the application of the above discussed methodology to the
entire dataset (\(6.1\) million of one-minute GHI measurements),
\(1.76\,\%\) were identified as CE events and \(0.036\,\%\) as ECE
events. The highest recorded GHI due to CE was
\(1416.6\,\text{W}/\text{m}^2\) on 24~May 2007 at a SZA of
\(19.9^\circ\) corresponding to OI of \(345.9\,\text{W}/\text{m}^2\) or
\(32.3\,\%\) above the threshold. The strongest CE event of \(49.7\,\%\)
above the clear sky threshold was observed on 28~October 2016 at a SZA
of \(59.2^\circ\) with a GHI value of \(861.8\,\text{W}/\text{m}^2\) and
a OI of \(286.1\,\text{W}/\text{m}^2\). Both cases are ECE events with
\(168.7\,\text{W}/\text{m}^2\) and \(155.4\,\text{W}/\text{m}^2\) above
the TSI at TSI for the same SZA, respectively. In the following sections
we are discussing the long-term trends and variability of the CE events
as well as of the corresponding OI and excess irradiation.

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the
yearly number of CE cases (each with duration of one minute), the yearly
mean OI and the yearly excess irradiation for the period 1993 -- 2023,
together with corresponding linear trends. All three quantities show
increasing trends, most pronounced for the frequency of occurrence
(\(+45.6\pm 21.9\,\text{cases}/\text{year}\)) and the excess irradiation
(\(+116.9 \pm 67.8\,\text{kJ}/\text{year}\)), which are also
statistically significant. In contrast the trend of the yearly mean OI
is negligible (\(+0\pm 0.1\,\text{W}/\text{m}^2/\text{year}\)) and of no
statistical significance. The average OI for the entire period is
\(+39.9\pm 2.7\,\text{W}/\text{m}^2\). The interannual variability of
the data about the trend lines is quite large. Furthermore, the spread
tends to increase with time (at least for the quantities of panels b and
c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

The distribution of the number of CE events as a function of OI is shown
in Figure\nobreakspace{}\ref{fig:ovir-distribution}. There is an inverse
relationship between the frequency of CE events and OI with an
exponential-like decline. This is expected, as the stronger the CE
events are, the rarer the conditions favoring the occurrence of CE
events. For the majority (over \(62\,\%\)) of the CE events the OI is
below the long-term average of \(39.9\,\text{W}/\text{m}^2\), while
about \(8.1\,\%\) of the events correspond to OI larger than
\(100\,\text{W}/\text{m}^2\) and up to the highest value of
\(412.4\,\text{W}/\text{m}^2\). This distribution is indicative of the
magnitude and the probability of the expected CE events over
Thessaloniki. Similar distribution of CE events, albeit with larger OI
values, has been reported by \citet{Vamvakas2020}, for the city of
Patras. This site is located \(2.5^\circ\) south of Thessaloniki and is
exposed to air masses coming mainly from the eastern Mediterranean
resulting in different cloud patterns, that may affect the
characteristics and magnitude of the CE events.

@@

The duration of the CE events is variable and can last for several
minutes or even more than an hour. In order to study the characteristics
of these consecutive events, we grouped them into bins of increasing
duration in steps of one minute. We have identified 28468 groups of CE
in the whole period of study, where the group of the longest duration of
140 minutes occurred on 07~July 2013.
Figure\nobreakspace{}\ref{fig:ceg-duration-distribution} shows the
frequency distribution of the CE events according to their duration. We
conclude that although some groups of events last for more than an hour,
about \(80\,\%\) have duration of less than 5 minutes.

>>>>

The data used in this study were recorded at the monitoring site of the
Laboratory of Atmospheric Physics, Aristotle University of Thessaloniki,
in Thessaloniki, Greece (\(40^\circ\,38'\,\)N, \(22^\circ\,57'\,\)E,
\(80\,\)m~a.s.l.). The GHI data were measured with a Kipp~\& Zonen CM-21
pyranometer and cover the period 01~January 1994 to 31~December 2023.
During the study period, the pyranometer was independently calibrated
three times at the Meteorologisches Observatorium Lindenberg, DWD,
verifying that the stability of the instrument's sensitivity was better
than \(0.7\,\%\) relative to the initial calibration by the
manufacturer. For the acquisition of radiometric data, the signal of the
pyranometer was sampled at a rate of \(1\,\text{Hz}\) with the mean and
standard deviation of these samples calculated and recorded every
minute. The measurements were corrected for the zero offset (``dark
signal'' in volts), which was calculated by averaging all measurements
recorded for a period of \(3\,\text{h}\), before (morning) or after
(evening) the Sun reaches an elevation angle of \(-10^\circ\). The
signal was converted to irradiance using the ramped value of the
instrument's sensitivity between subsequent calibrations.

@@

Following the application of the above discussed methodology to the
entire dataset (\(6\) million of one-minute GHI measurements),
\(1.78\,\%\) were identified as CE events and \(0.037\,\%\) as ECE
events. The highest recorded GHI due to CE was
\(1416.6\,\text{W}/\text{m}^2\) on 24~May 2007 at a SZA of
\(19.9^\circ\) corresponding to OI of \(345.9\,\text{W}/\text{m}^2\) or
\(32.3\,\%\) above the threshold. The strongest CE event of \(49.7\,\%\)
above the clear sky threshold was observed on 28~October 2016 at a SZA
of \(59.2^\circ\) with a GHI value of \(861.8\,\text{W}/\text{m}^2\) and
a OI of \(286.1\,\text{W}/\text{m}^2\). Both cases are ECE events with
\(168.7\,\text{W}/\text{m}^2\) and \(155.4\,\text{W}/\text{m}^2\) above
the TSI at TSI for the same SZA, respectively. In the following sections
we are discussing the long-term trends and variability of the CE events
as well as of the corresponding OI and excess irradiation.

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the
yearly number of CE cases (each with duration of one minute), the yearly
mean OI and the yearly excess irradiation for the period 1993 -- 2023,
together with corresponding linear trends. All three quantities show
increasing trends, most pronounced for the frequency of occurrence
(\(+45.6\pm 21.9\,\text{cases}/\text{year}\)) and the excess irradiation
(\(+116.9 \pm 67.8\,\text{kJ}/\text{year}\)), which are also
statistically significant. In contrast the trend of the yearly mean OI
is negligible (\(+0\pm 0.1\,\text{W}/\text{m}^2/\text{year}\)) and of no
statistical significance. The average OI for the entire period is
\(+40.1\pm 2.5\,\text{W}/\text{m}^2\). The interannual variability of
the data about the trend lines is quite large. Furthermore, the spread
tends to increase with time (at least for the quantities of panels b and
c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

The distribution of the number of CE events as a function of OI is shown
in Figure\nobreakspace{}\ref{fig:ovir-distribution}. There is an inverse
relationship between the frequency of CE events and OI with an
exponential-like decline. This is expected, as the stronger the CE
events are, the rarer the conditions favoring the occurrence of CE
events. For the majority (over \(62\,\%\)) of the CE events the OI is
below the long-term average of \(40.1\,\text{W}/\text{m}^2\), while
about \(8.1\,\%\) of the events correspond to OI larger than
\(100\,\text{W}/\text{m}^2\) and up to the highest value of
\(412.4\,\text{W}/\text{m}^2\). This distribution is indicative of the
magnitude and the probability of the expected CE events over
Thessaloniki. Similar distribution of CE events, albeit with larger OI
values, has been reported by \citet{Vamvakas2020}, for the city of
Patras. This site is located \(2.5^\circ\) south of Thessaloniki and is
exposed to air masses coming mainly from the eastern Mediterranean
resulting in different cloud patterns, that may affect the
characteristics and magnitude of the CE events.

@@

The duration of the CE events is variable and can last for several
minutes or even more than an hour. In order to study the characteristics
of these consecutive events, we grouped them into bins of increasing
duration in steps of one minute. We have identified 28062 groups of CE
in the whole period of study, where the group of the longest duration of
140 minutes occurred on 07~July 2013.
Figure\nobreakspace{}\ref{fig:ceg-duration-distribution} shows the
frequency distribution of the CE events according to their duration. We
conclude that although some groups of events last for more than an hour,
about \(80\,\%\) have duration of less than 5 minutes.
